---
title: CV
name: Sydney
surname: Rilum
position: "Master's Student"
address: "Bren School of Environmental Science & Management, UC Santa Barbara"
email: "sydneyrilum@ucsb.edu"
github: sydneyrilum
linkedin: sydneyrilum
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  vitae::markdowncv:
    theme: davewhipp
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(vitae)
```

## Education

```{r}
library(tibble)
tribble(
  ~ Institution, ~ Year, ~ Where, ~ Degree,
  "Bren School of Environmental Science & Management, UC Santa Barbara", "Expected 2022", "Santa Barbara, CA", "Master of Environmental Science and Management",
  "University of San Diego", "2014 - 2018", "San Diego, CA", "Bachelor of Arts in Environmental and Ocean Sciences",
  ) %>% 
  detailed_entries(Institution, Year, Where, Degree)
```

## Work Experience

```{r}
library(tibble)
tribble(
  ~ Position, ~ Year, ~ Where, ~ Institution,
  "California Department of Fish and Wildlife", "5/2020 - 6/2020", "Newport Beach, CA", "Scientific Aid,",
  "Laguna Ocean Foundation", "5/2019 - 9/2020", "Laguna Beach, CA", "Tidepool Educator,",
  "Duke University Marine Laboratory", "6/2017 - 8/2017", "Beaufort, NC", "NSF-Funded Research Intern,",
  "Aquarium of the Pacific", "5/2016 - 8/2016", "Long Beach, CA", "Marine Mammalogy Intern,",
  ) %>% 
  detailed_entries(Position, Year, Where, Institution)
```

## Volunteer Experience

```{r}
library(tibble)
tribble(
  ~ Position, ~ Year, ~ Where, ~ Organization,
  "California Sea Grant - CoAST SB", "10/2020 - Present", "Santa Barbara, CA", "Beach Profile Monitor,",
  "OC Habitats", "10/2018 - 9/2020", "Orange County, CA", "Shorebird Field Monitor & Education Outreach Volunteer,",
  "Ocean Discovery Institute", "1/2018 - 3/2018", "San Diego, CA", "Science Education Mentor,",
  ) %>% 
  detailed_entries(Position, Year, Where, Organization)
```

## Conferences

```{r}
library(tibble)
tribble(
  ~ Research, ~ Year, ~ Where, ~ Conference,
  "Awarded 2nd place for undergraduate research presentation on the Effect of Adult Chemical Cues on Molting of Fiddler Crab Megalopae in Low Salinity Seawater", "4/4/2019", "St. John's, Newfoundland, Canada", "48th Benthic Ecology Meeting",
  "Effect of Adult Chemical Cues on Molting of Fiddler Crab Megalopae in Low Salinity Seawater", "4/19/2018", "San Diego, California", "Creative Collaborations Undergraduate Research Conference",
  ) %>% 
  detailed_entries(Research, Year, Where, Conference)
```

## Publication

**Rilum, S.C. (2018).** Effect of Adult Chemical Cues on Molting of Fiddler Crab Megalopae in Low Salinity Seawater. *Undergraduate Honors Theses*, 58. [http://digital.sandiego.edu/honors_theses/58](http://digital.sandiego.edu/honors_theses/58)

## Skills & Certifications

**Computer:** R/R Studio, ArcGIS/QGIS (basic mapping), Microsoft Office (Word, Excel, Powerpoint, Outlook)

**Languages:** English, Conversational Spanish, Basic Danish

**Certifications:** PADI Open Water Diver, California Boating License



```{r, include = FALSE}
## Awards
tribble(
  ~Year, ~Type, ~Desc,
  2020, "Physics", "Awarded for her work on radioactivity with Pierre Curie and Henri Becquerel",
  1911, "Chemistry", "Awarded for the discovery of radium and polonium"
) %>% 
  brief_entries(
    glue::glue("Nobel Prize in {Type}"),
    Year, 
    Desc
  )
```

```{r, include = FALSE}
## Publications
library(dplyr)
knitr::write_bib("vitae", "packages.bib")

bibliography_entries("packages.bib") %>%
  arrange(desc(author$family), issued)
```

