---
title: 'Two-sample t-test'
description: |
  Comparison of female lizard lengths for two species
author:
  - name: Sydney Rilum
date: 12-03-2020
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(janitor)
library(ggbeeswarm)
library(kableExtra)
library(effsize)
library(broom)
```

## Introduction 

This report provides an exploratory comparison between the total lengths of two species of female lizards, the Desert Grassland lizard (*Cnemidophorus uniparens*) and the Colorado Checkered lizard (*Cnemidophorus tessalatus*), using the Jornada Basin LTER lizards data set (Lightfoot 2017). Total female lizard length data will be visually explored as well as statistically analyzed through a two-sample t-test to determine whether there is a significant difference in mean total length for female lizards of each species. 

```{r}
# Read in the data and clean column names
lizard <- read.csv(here("data", "lizards.csv")) %>% 
  clean_names()
```

## Lizard Length Comparison Results and Analysis

```{r}
# Create a new subset of data
lizard_length <- lizard %>% 
  select(spp, sex, total_length) %>%  # selects three variables
  filter(sex == "F", # filters for female lizards
         spp %in% c("CNUN", "CNTE")) %>%  # filters to only include observations for CNUN and CNTE species
  mutate(spp = case_when(
    spp == "CNUN" ~ "Desert Grassland", 
    spp == "CNTE" ~ "Colorado Checkered")) 

# Create a finalized visualization comparing total lengths of female lizards for the two species.
ggplot(data = lizard_length, 
       aes(x = spp, y = total_length)) +
  geom_beeswarm(aes(color = spp,
                    alpha = 0.5)) +
  scale_color_manual(values = c("darkorange","springgreen4")) +
  geom_boxplot(fill = NA, 
               width = 0.1) +
  stat_summary(fun=mean, 
               geom="point", 
               shape=20, 
               size=3, 
               color="black", 
               fill="black") +
  labs(x = "Species",
       y = "Total Length (mm)",
       title = "Total Lengths of Female Lizards") +
  theme_light() +
  theme(legend.position = "none")
```

**Figure 1.** Comparison between the total lengths (mm) for female lizards of the Colorado Checkered (*Cnemidophorus tessalatus*) and Desert Grassland (*Cnemidophorus uniparens*) species. Box endpoints indicate the 25^th^ and 75^th^ percentile values; the black line and black point within the box indicate the median and mean value for each species, respectively. (Data: Lightfoot 2017).

<br>

**Table 1.** Descriptive statistics (mean, standard deviation, and sample size) for total length of female lizards for the two species: Colorado Checkered (*Cnemidophorus tessalatus*) and Desert Grassland (*Cnemidophorus uniparens*). (Data: Lightfoot 2017).

```{r}
# Descriptive statistics data subset
lizard_length_stats <- lizard_length %>% 
  select(spp, total_length) %>% 
  group_by(spp) %>% 
  summarize(mean_length = round(mean(total_length, na.rm = TRUE),2),
            sd_length = round(sd(total_length, na.rm = TRUE),2),
            sample_size = n())

# Finalized summary table 
lizard_length_stats %>% 
  kable(col.names = c("Species", "Mean Length (mm)", "Standard deviation (mm)", "Sample size")) %>% 
  kable_styling()
```

```{r, include = FALSE}
## T-test preparation code: 

# Create vectors of lizard length for each species 
lizard_colorado <- lizard_length %>% 
  filter(spp == "Colorado Checkered") %>% 
  pull(total_length)

lizard_desert <- lizard_length %>% 
  filter(spp == "Desert Grassland") %>% 
  pull(total_length)

# Plot histograms & qq-plots to check assumptions
hist(lizard_colorado) # Looks normal
qqnorm(lizard_colorado) # Looks relatively linear
hist(lizard_desert) # Looks normal, possibly a bit negatively skewed
qqnorm(lizard_desert) # Looks linear

# Assumptions of normality hold, & sample size additionally > Central Limit Theorem (means will be normally distributed). Therefore, it is okay to use a two-sample t-test for means comparison! 

## T-test: Is there a significant difference in total length difference in mean total length between female lizards for the two species? 

# Run the t-test
lizard_ttest <- t.test(lizard_colorado, lizard_desert)
lizard_ttest

# Get the tidy model results to call outputs in-line
lizard_ttest_tidy <- tidy(lizard_ttest)

# Get the means & SDs for each, removing NA values
colorado_mean <- mean(lizard_colorado)
colorado_sd <- sd(lizard_colorado)
desert_mean <- mean(lizard_desert, na.rm = TRUE)
desert_sd <- sd(lizard_desert, na.rm = TRUE)

# Find the effect size (Cohen’s d) of the difference between mean lengths for the two species.
lizard_effsize <- cohen.d(lizard_colorado, lizard_desert, na.rm = TRUE)
lizard_effsize
```

On average, female Colorado Checkered lizards are longer than female Desert Grassland lizards (`r round(colorado_mean,2)` $\pm$ `r round(colorado_sd, 2)` and `r round(desert_mean,2)` $\pm$ `r round(desert_sd, 2)` mm, respectively; mean $\pm$ 1 standard deviation) (Figure 1, Table 1). While the absolute difference in means is `r round(colorado_mean, 2) - round(desert_mean, 2)` mm (a `r round((colorado_mean - desert_mean)/((colorado_mean + desert_mean)/2)*100, 2)`% difference), the difference in means is significant (Welch's two-sample t-test: t(`r round(lizard_ttest_tidy$parameter,2)`) = `r round(lizard_ttest_tidy$statistic,2)`, p < 0.001), and the effect size is large (Cohen's *d* = `r round(lizard_effsize$estimate,2)`). 

<br>

### Citation:

[Lightfoot, D. 2017. Lizard pitfall trap data (LTER-II, LTER-III) ver 36. Environmental Data Initiative.](https://doi.org/10.6073/pasta/ff37b4cace16a9943575f3fd7067064e)
