---
title: "Working with Raster Data"
description: |
  Raster of cetacean species richness off the California coast
author:
  - name: Sydney Rilum
date: 02-21-2021
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(raster) # attach this BEFORE tidyverse
library(tidyverse)
library(here)
library(sf)
library(fasterize)
library(rnaturalearth)
```

```{r}
## Create a raster of cetacean species richness off the California coast

# Read in cetaceans shapefiles all together
cetaceans_files <- list.files(path = here("data", "ca_cetaceans"), full.names = TRUE)

# Rasterize all .tif files at once
cetaceans_stack <- raster::stack(cetaceans_files) 
  
# Convert into a data frame
cetaceans_df <- rasterToPoints(cetaceans_stack) %>%  # convert raster to points
  as_tibble()  # convert points to a data frame

# Wrangle the data to determine number of species present in each (x,y) cell (i.e. species richness)
species_richness <- cetaceans_df %>% 
  pivot_longer(cols = 3:37,
               names_to = "species",
               values_to = "probability") %>% 
  mutate(presence = case_when(  # apply probability threshold of 0.6
    probability >= 0.6 ~ 1,  # 1 = species present
    probability < 0.6 ~ 0)) %>%  # 0 = species not present
  filter(presence == 1) %>% 
  group_by(x, y) %>%  # group by coordinates/cells
  count(presence)
```

```{r, results = "hide"}
# Read in California coastline shapefile from `rnaturalearth` package
california <- ne_download(scale = 110, type = 'coastline', category = 'physical', returnclass = "sf") 
```

### Raster of Cetacean Species Richness off the California Coast

```{r}
# Create a map of cetacean species richness off the California coast
ggplot() +
  geom_raster(data = species_richness, 
              aes(x = x, y = y, fill = n)) +
  geom_sf(data = california) + # shapefile of the coastline
  coord_sf(expand = 0,  # define boundaries of map to match the raster extent
           xlim = c(-125, -115), 
           ylim = c(32, 38)) +
  scale_fill_gradientn(colors = c("lightgray", "steelblue1","darkblue")) +
  theme_light() +
  theme(panel.background = element_rect(fill = "white")) +
  labs(fill = "Number of \nCetacean Species",
       x = "Longitude",
       y = "Latitude")
```

**Figure 1.** A map depicting a raster of species richness for 35 cetacean species off the coast of California. Raster of species richness was created by considering a species as "present" in each cell if its probability of occurrence was greater than or equal to a threshold value of 0.6. Darker blue cells indicate areas with higher species richness. (Data: Kaschner *et al.* 2016).

<br>

### Citations:

Cetacean raster data: [Kaschner, K., Rius-Barile, J., Kesner-Reyes, K., Garilao, C., Kullander, S., Rees, T., & Froese, R. (2016). AquaMaps: Predicted range maps for aquatic species.](https://www.aquamaps.org/)

California coastline layer: [Andy South (2017). rnaturalearth: World Map Data from Natural Earth. R package version 0.1.0.](https://CRAN.R-project.org/package=rnaturalearth)
